Adaptive evolution with aneuploidy and mutation

### 2022-04-18 ms/ms.tex
moved tables with fitness values inside the model illustration figure; also added table for s,b,c calculation

%%%%%%%% Fig illustation
\begin{figure}[b!]
  \centering
  \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
  \label{fig:model1}
  \caption{
    \textbf{Model illustration.}
    There are four genotypes in our model: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt\; to \eumt.
    Arrows denote transitions between genotypes, with transitions rates: $\mu$, beneficial mutation rate; $\delta$, aneuploidy rate.
  }
  \label{fig:models}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%

@@

Transitions between the genotypes occur as follows~(\Cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \Cref{table:fitness}.

%%% Table: fitness
\begin{table}[h]
\centering
\caption{\textbf{Fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & \euwt & \anwt & \anmt & \eumt \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
\end{tabular}
\label{table:fitness}
\caption*{
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \Cref{table:fitness} and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets.
%% estimates generated by MAP-and-hdi.ipynb
The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimates  $6.7\cdot10^{-6}$ aneuploidy gain rate, \citet{Kumaran2013} estimates $3.0-4.3\cdot10^{-5}$ aneuploidy loss rate (95\% confidence interval).
The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
Thus, we can infer that the benefit and cost of trisomy was $b=2.5\%$ and $c=0.3\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{table:fitness}).

>>>>

%%%%%%%% Fig illustation
\begin{figure}[b!]
  \centering

  \begin{subfigure}[t]{0.03\textwidth} \textbf{A} \end{subfigure}
  \begin{subfigure}{0.75\textwidth}
    \centering
    \includegraphics[height=1.8in]{../figures/Fig1-A.pdf}      
  \end{subfigure}

  \begin{subfigure}[t]{0.03\textwidth} \textbf{B} \end{subfigure}
  \begin{subfigure}{0.75\textwidth}
	\centering
	\begin{tabular}{lllll}
	\emph{Genotype} $i$ & \euwt & \anwt & \anmt & \eumt \\
	\hline
	\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)(1+s)+b$ & $1+s$               
	\end{tabular}
	\label{table:fitness}
  \end{subfigure}

    \begin{subfigure}[t]{0.03\textwidth} \textbf{C} \end{subfigure}
    \begin{subfigure}{0.75\textwidth}
	\centering
	\begin{tabular}{llll}
	\emph{s, b, c} & s & c & b \\
	\hline
	 \emph{Calculation} & $w_{\eumt-1}$ & $1-\dfrac{w_\anmt-w_\anwt}{w_\eumt-1}$ & $w_\anwt - 1 + c$               
	\end{tabular}
	\label{table:fitness}
  \end{subfigure}

  \caption{
    \textbf{The model.} 
    \textbf{(A)} Model Illustration.
    There are four genotypes in our model: euploid wild-type, \euwt; euploid mutant, \eumt; aneuploid wild-type, \anwt; and aneuploid mutant, \anmt.
    Overall there are two possible trajectories from \euwt\; to \eumt.
    Arrows denote transitions between genotypes, with transitions rates: $\mu$, beneficial mutation rate; $\delta$, aneuploidy rate.
    \textbf{(B)} Fitness values. $s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
   \textbf{(C)} Calculations of $s,b,c$ from $w_i$ values, derived from (B) expressions.
   }
  \label{fig:models}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%

@@

Transitions between the genotypes occur as follows~(\Cref{fig:models}A): Beneficial mutations from \euwt\; to \eumt\; and from \anwt\; to \anmt\; occur with probability $\mu$, the mutation rate. We neglect back-mutations (i.e., from \eumt\; to \euwt\; and from \anmt\; to \anwt).
Aneuploidy is formed by chromosome mis-segregation, so that cells transition from \euwt\; to \anwt\; and from \anmt\ to \eumt\ with probability $\delta$, the aneuploidy rate. That is, we assume chromosomes are gained and lost at the same rate, and we neglect events that form a less-fit genotype (i.e., \anwt\; to \euwt\; and \eumt\; to \anmt).
The fitness values of the four genotypes are given by \Cref{fig:models}B.

The initial population has $N$ cells with genotype \euwt.
The effect of natural selection on the frequency $f_i$ of genotype $i = 2n, 2n+1, 2n+1^*, \text{or } 2n^*$ is given by
    \begin{equation} \label{eq:selection-single} 
      f^s_i = \frac{f_i w_i}{\bar{w}} \;,
    \end{equation}
where the fitness values $w_i$ are given in \Cref{fig:models}B and $\bar{w} = \sum_{j}{f_j w_j}$ is the population mean fitness.
The effect of mutation and aneuploidy on genotype frequencies is given by
    \begin{equation} \label{eq:mutation-aneuploidy-single}
    \begin{aligned}
      &f^m_{2n} &=&\; (1 - \delta - \mu) f^s_{2n}  \;,\\
      &f^m_{2n+1} &=&\; \delta f^s_{2n} + (1 - \mu) f^s_{2n+1}  \;,\\
      &f^m_{2n+1^*} &=&\; \mu f^s_{2n+1} + (1-\delta) f^s_{2n+1^*}  \;,\\
      &f^m_{2n^*} &=&\; \mu f^s_{2n} + \delta f^s_{2n+1} + f^s_{2n^*}  \;.
    \end{aligned}
    \end{equation}
Finally, random genetic drift is modeled using a multinomial distribution~\citep{Otto2007},
    \begin{equation} \label{eq:drift-single}
      \vec{f'} \sim \frac{1}{N} \cdot \mathit{Mult}(N,\ \vec{f^m}) \;,
    \end{equation}
where $\vec{f^m}=(f^m_{2n}\ ,\ f^m_{2n+1}\ ,\ f^m_{2n+1^*}\ ,\ f^m_{2n^*})$ are the frequencies of the genotypes after mutation and aneuploidy, $\vec{f'}$ are the genotype frequencies in the next generation, and $Mult(N,\ \vec{f})$ is a multinomial distribution parameterized by the population size $N$ and the genotype frequencies $\vec{f}$.
Overall, the change in genotype frequencies from one generation to the next is given by the transformation $f_i \to f'_i$.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of model parameters~(\Cref{fig:posterior}). 
We report parameter estimates using the MAP (maximum a posteriori) and providing the 50\% HDI (highest density interval) in square brackets.
%% estimates generated by MAP-and-hdi.ipynb
The estimated mutation rate, $\mu=2.942\cdot10^{-6}\ [2.017\cdot10^{-7}-4.15\cdot10^{-6}]$, corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014} or $3.3\cdot10^{-10}$~\citep{Lynch2008}.
The estimated aneuploidy rate, $\delta=1.722\cdot10^{-3}\ [1.394\cdot10^{-3}-2.754\cdot10^{-3}]$ is higher than in previous studies: for chromosome III in diploid \yeast, \citet{Zhu2014} estimates  $6.7\cdot10^{-6}$ aneuploidy gain rate, \citet{Kumaran2013} estimates $3.0-4.3\cdot10^{-5}$ aneuploidy loss rate (95\% confidence interval).
The estimated fitness values are $w_{\anwt}=1.022\ [1.021-1.023]$,
$w_{\anmt}=1.025\ [1.024-1.026]$,
$w_{\eumt}=1.028\ [1.026-1.029]$, all relative to the fitness of \euwt, which is set to $w_{\euwt}=1$. 
Thus, we can infer that the benefit and cost of trisomy was $b=92\%$ and $c=89\%$, and the benefit of the beneficial mutation was $2.8\%$~(\Cref{fig:models}C).
